Relational Parametricity for Algebraically Indexed Types

### 2012-06-12 paper/section-general-framework.tex
Add existential types, and extend the proof of the substitution lemma

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i : s \vdash A \isType}
    {\Delta \vdash \forall i \mathord: s. A \isType}
  \end{mathpar}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some sort.

@@

  \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
\end{mathpar}
\begin{lemma}
  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
  \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
    \isType$.
  \end{enumerate}
\end{lemma}

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. Accordingly, we assume that each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types are simply
the corresponding constructions on sets. The interpretation of the
universal quantifier ignores the indexing: the interpretation of the
type $\forall i\mathord:s.A$ is exactly the interpretation of the type
$A$.

@@

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be
$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s}
= \rho \}$. The set of extensions of a relational environment will be
used in the interpretation of the universal quantification type.

We assign a relational interpretation to all well-indexed types
$\Delta \vdash A \isType$ by induction on their derivations,
parameterised by relational environments $\rho \in
\relEnv{E}(\Delta)$:
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

>>>>

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \forall i \mathord: s. A \isType}

    \inferrule* [right=TyExist]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types are constructed using the \TirName{TyExist}
rule.

@@

  \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A

  \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
\end{mathpar}
\begin{lemma}
  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
  \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
    \isType$.
  \end{enumerate}
\end{lemma}

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. Accordingly, we assume that each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types are simply
the corresponding constructions on sets. The interpretations of the
universal and existential quantifiers ignore the indexing: the
interpretation of the type $\forall i\mathord:s.A$ is exactly the
interpretation of the type $A$, and likewise for $\exists
i\mathord:s.A$.

@@

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be
$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s}
= \rho \}$. The set of extensions of a relational environment will be
used in the interpretation of the universal and existential type
formers.

We assign a relational interpretation to all well-indexed types
$\Delta \vdash A \isType$ by induction on their derivations,
parameterised by relational environments $\rho \in
\relEnv{E}(\Delta)$:
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \} \\
  \rsem{\exists i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcup\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.
